### 2012-07-27 Doc/VRIPHYS2012/vriphys2012.tex
article
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@433 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\title[Short title]%
      {Bezier shell finite element for interactive surgical simulation} % TODO

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
\author[...]
       {...} % TODO

@@

This problem has been faced by the computer graphics community but for different applications.
For instance, many related work addresses the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003,Choi2007}. %% Todo verify.... 
However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
This is completely acceptable for some computer graphics application, where the parameters can be tuned manually to obtain a realistic animation.
In the case of medical simulation, more physics-based modeling is needed.

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.  %% Do we
While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling of deformations \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory was integrated numerically \CD{Todo citation of a shell book} using Finite Elements, but there are two issues for obtaining interactive frame rates: 
(i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and 
(ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
This type of interpolation is necessary because bending continuum model
(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
Ubach and al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

@@

\begin{equation}
    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}.
\end{equation}

@@

\begin{equation}
    \begin{split}
        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
                   & + \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_j + \\
                   & + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \mathbf{\theta}_i
    \end{split}
\end{equation}

@@

\begin{equation}\label{eq-cauchy}
    J_m = \left[ \begin{matrix}
        \deriv{u_x}{x} \\
        \deriv{u_y}{y} \\
        \deriv{u_x}{y} + \deriv{u_y}{x}
    \end{matrix} \right]
\end{equation}
.

@@

Because the deformation field for the shell is non-linear \eqref{eq-bezU}
the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
numerical integration. In our implementation we employ 6-point Gaussian
quadrature for integration over triangle area.
% TODO: add ref
% TODO: add abscissas/weights?

@@

%The displacement field defined in \eqref{eq-bezU} is clearly symmetric in
%respect to deformations, but ...

@@

In next two subsections we present tests to validate the convergence of our
B\'ezier shell element. The results are compared with the shell element
composed of DKT element and optimal ANDES element \cite{Felippa2003}. The
last subsection presents evaluation of the computational complexity of our
solution.

@@

\subsection{Applications}

The presented element has been successfully implemented in SOFA framework
\cite{Faure2012} for real-time surgical simulations. It has been evaluated
on several applications.

% TODO: !!! When ANONYMISING do something with the citation !!!
One being a tool for pre-operative planning of corrective surgery of
congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
surgery is a complicated process where several arterial vessels have to be
cut, patched and reconnected. The element has been used for to model
deformations of blood vessels after suturing and/or applying a artificial
patch. Figure \ref{fig-chd} presents a screenshots of the application
prototype.

@@

Another designed application is simulation of women reproductive organs
(vagaina/uterus/\ldots). 
The simulation is so as to evaluate
stress on a tissue during a birth.

@@

This paper presents new shell element with geometry and displacement
function based on the formulation of B\'ezier triangle. It fixes problems
of the previous model \cite{Comas2010c}, namely:

@@

    \item It is continuous on the edges and provides $C^1$ continuity in
    corners of the element. This is essential for good visual experience if
    high-resolution mesh is mapped on the mechanical mesh.

@@

    \item TODO: anything else \ldots? % TODO

@@

>>>>

\title[B\'ezier Shell FE]%
      {B\'ezier Shell Finite Element for Interactive Surgical Simulation}

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
% TODO: anonymize
\author[...]
       {...} % TODO

@@

This problem has been faced by the computer graphics community but for different applications.
For instance, many related works address the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003}. %% Todo Choi2007?
However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
This is completely acceptable for some computer graphics application, where the parameters can be tuned manually to obtain a realistic animation.
In the case of medical simulation, more physics-based modeling is needed.

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in real-time simulations ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling of deformations \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory was integrated numerically \CD{Todo citation of a shell book} using Finite Elements, but there are two issues for obtaining interactive frame rates: 
(i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and 
(ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaining a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynomials.
This type of interpolation is necessary because bending continuum model
(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
On the other hand, elements common in mechanics (e.g. Discrete Kirchhoff Triangles) are based on shape functions that are impractical for real-time simulations or don't have explicitly defined shape functions at all.
In \cite{Comas2010c}, Comas et al. use a hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
Ubach et al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

@@

\begin{equation}
    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}
\end{equation}

@@

\begin{equation}\label{bez-delta}
    \begin{split}
        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
                   & + \sum_{j=4}^9 P_{i-j} \times \delta \mathbf{\theta}_j + \\
                   & + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \mathbf{\theta}_i
    \end{split}
\end{equation}

@@

\begin{equation}\label{eq-cauchy}
    J_m = \left[ \begin{matrix}
        \deriv{u_x}{x} \\
        \deriv{u_y}{y} \\
        \deriv{u_x}{y} + \deriv{u_y}{x}
    \end{matrix} \right]
\end{equation}

@@

Because the deflection field for the element is non-linear \eqref{eq-delta}
in position the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
numerical integration. In our implementation we employ 6-point Gaussian
quadrature for integration over triangle area.
% TODO: add ref
% TODO: add abscissas/weights?

@@

%The displacement field defined in \eqref{eq-delta} is clearly symmetric in
%respect to deformations, but ...

@@

In next two subsections we present tests to validate the convergence of our
B\'ezier shell element. We use a combination of two elements that are very well
established in the area of mechanical modelling.
Our results are compared with the shell element
composed of Discrete Kirchoff Triangle (DKT) element \cite{Batoz1980} for bending plate and optimal ANDES element \cite{Felippa2003} for membrane.
The last subsection presents evaluation of the computational complexity of our solution.

@@

\subsection{Application on Congenital Heart Disease Correction Planning}

Surgical interventions in infants with congenitally malformed great arteries
and hearts are extremely challenging due to the complex and heterogeneous
nature of their disease patterns. At present, cardiac surgeons rely on
non-invasive imaging for patient-specific examination and a set of preoperative
sketches with varying approaches to the surgical procedure. However, the most
promising approach is often chosen during the actual open-heart surgery, when
surgeons get a more concrete idea of possible outcomes. Finalizing this
decision during an on-pump cardiac surgery is further complicated by deflated
blood vessels since it is difficult to imagine their exact change of shape
caused by the intervention. The important decisions have to be made in very
short time and are strongly dependent on the surgeon's experience.

The planning process can be improved by employing surgical simulation system
using three-dimensional heart models based on patient-specific images as input.
Most present simulation systems focus on execution of low-level surgical
procedures, e.g. incising and suturing. In these reactive simulation systems it
is more important to enable simulations in real-time than accurate simulation
of tissue and blood vessel deformation in response to simulated interventions.

To be able to predict results of complex surgical procedures a physically more
accurate approach is necessary. Using our shell finite element model we have
developed a prototype \cite{Kislinskiy2012} of new surgical simulation system
for preoperative planning. The element has been used to model deformations of
blood vessels and artificial patch. By performing a set of topological changes
we simulate the process of incising the vessel and suturing on a pre-specified
boundary. By iteratively adjusting the underlaying mesh we search for the
elastic equilibrium. Figure \ref{fig-chd} presents a screenshots of the
application prototype.

%%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%
%% TODO: is this OK for anonymity?
%The presented element has been successfully implemented in SOFA framework
%\cite{Faure2012} for real-time surgical simulations. It has been evaluated
%on several applications.
%
%% TODO: !!! When ANONYMISING do something with the citation !!!
%One being a tool for pre-operative planning of corrective surgery of
%congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
%surgery is a complicated process where several arterial vessels have to be
%cut, patched and reconnected. The element has been used for to model
%deformations of blood vessels after suturing and/or applying a artificial
%patch. Figure \ref{fig-chd} presents a screenshots of the application
%prototype.

@@

% TODO?
%Another designed application is simulation of women reproductive organs
%(vagaina/uterus/\ldots). 
%The simulation is so as to evaluate
%stress on a tissue during a birth.

@@

In this paper we presents new shell element with interpolation functions
based on B\'ezier triangle which is more suitable for real-time simulations than elements common in mechanics with complex, obscure or unclear interpolation functions.
It fixes problems of the previous model \cite{Comas2010c} based on cubic polynomials, namely:

@@

    \item Element is based on all 6 DOFs, including the drilling degree of
        freedom, not just 5 DOFs. Therefore it doesn't have the problem of
        undefined z-rotations.

@@

    \item It is $C^0$ continuous on the edges and provides $C^1$ continuity in
    corners of the element. This is essential for good visual experience if
    high-resolution mesh is mapped on the mechanical mesh.

@@

Unified formulation of the deformation allows for derivation of three-dimensional 
stress tensor. This however remains as topic for future work and
we have kept the computation of membrane stresses separate from bending stresses.
